---
title: "Lab 11 - Grading the professor, Pt. 2"
author: Tsion
output: github_document
---

## Load packages and data

```{r load-packages, message=FALSE}
library(tidyverse) 
library(openintro)
library(titanic)
library(readxl)
```

```{r load-data, message=FALSE}
# Load the Titanic 3 dataset
titanic3 <- read_excel(
  "data/titanic3.xls",
  col_types = c(
    "numeric", "numeric", "text", "text", "numeric", "numeric", "numeric",
    "text", "numeric", "text", "text", "text", "text", "text"
  )
)

# Load the training and test datasets from the titanic package
data("titanic_train")
data("titanic_test")
```

```{r clean-data}
# Rename columns to lowercase to make them easier to work with
names(titanic3) <- names(titanic3) %>% tolower()
names(titanic_train) <- names(titanic_train) %>% tolower()
names(titanic_test) <- names(titanic_test) %>% tolower()

# Convert passenger class to an ordered factor
# 3 = third class, 2 = second class, 1 = first class
titanic3 <- titanic3 %>%
  mutate(pclass_ord = factor(pclass, ordered = TRUE, levels = c(3, 2, 1)))

titanic_train <- titanic_train %>%
  mutate(pclass_ord = factor(pclass, ordered = TRUE, levels = c(3, 2, 1)))

titanic_test <- titanic_test %>%
  mutate(pclass_ord = factor(pclass, ordered = TRUE, levels = c(3, 2, 1)))

# Look at the full Titanic dataset
glimpse(titanic3)
summary(titanic3)
```

I can see that the full `titanic3` dataset has 1,309 passengers. There are also missing values in some variables, especially variables like `body`, which makes sense because not every body was recovered after the sinking.

### Part 1: Apparent accuracy

## Exercise 1

```{r exercise1_code}
# Fit a logistic regression model predicting survival from sex and passenger class
m_apparent <- glm(
  survived ~ sex + pclass,
  data = titanic3,
  family = binomial
)

# View model output
summary(m_apparent)
```

### 1.2

```{r predicted-probabilities}
# Generate predicted survival probabilities for each passenger
p_apparent <- predict(m_apparent, type = "response")

# Look at the first few predicted probabilities
head(p_apparent)
```

### 1.3

I used 0.5 as the cutoff of predicting survival and non-survival 

```{r class-predictions}
# Convert probabilities into predicted classes
# If probability is greater than 0.5, predict survived
# Otherwise, predict did not survive
yhat_apparent <- ifelse(p_apparent > 0.5, 1, 0)

# Look at the first few predicted classes
head(yhat_apparent)
```

### 1.4

```{r apparent-accuracy}
# Calculate apparent accuracy
acc_apparent <- mean(yhat_apparent == titanic3$survived, na.rm = TRUE)

acc_apparent
```

### 1.5

Apparent accuracy is likely to overestimate true predictive performance because the model is being tested on the same data it was trained on. 

In my opinion, this is like studying the answer key and then taking the same test. The model may look good because it already had access to the full dataset.

Another analogy I can think of is is practicing a speech in front of a mirror and then judging your public speaking ability only from that mirror practice. It tells you something, but it does not really show how well you would perform in a new situation.

### Part 2: Holding passengers back

## Exercise 2

## 2.1

```{r exercise2_code}
# Fit the same logistic regression model, but only using the training data
m_split <- glm(
  survived ~ sex + pclass,
  data = titanic_train,
  family = binomial
)

# View the model output
summary(m_split)
```

### 2.2

```{r training-accuracy}
# Generate predicted probabilities for the training data
p_train <- predict(m_split, type = "response")

# Convert probabilities to predicted survival decisions using 0.5 as the cutoff
yhat_train <- ifelse(p_train > 0.5, 1, 0)

# Calculate training accuracy
acc_train <- mean(yhat_train == titanic_train$survived, na.rm = TRUE)

acc_train
```

### 2.3

```{r test-accuracy}
# The Titanic test data does not include survived, so we join it from titanic3
# We use name and ticket to match passengers between the datasets
titanic_test_survival <- titanic_test %>%
  left_join(
    titanic3 %>% select(name, ticket, survived),
    by = c("name", "ticket")
  )

# Generate predicted probabilities for the test data
p_test <- predict(m_split, newdata = titanic_test_survival, type = "response")

# Convert probabilities to predicted survival decisions
yhat_test <- ifelse(p_test > 0.5, 1, 0)

# Calculate test accuracy
acc_test <- mean(yhat_test == titanic_test_survival$survived, na.rm = TRUE)

acc_test
```

Ok looks good!

### 2.4

The training accuracy is larger than the test accuracy. The training accuracy is about 0.787, while the test accuracy is about 0.761. This is the typical pattern because the model was fit using the training data, so it usually performs a little better on data it has already seen. The test data is more difficult because those passengers were held back and were not used to build the model.

For Lloyd’s, the test accuracy is closer to what they actually need. Lloyd’s would not just want to know how well the model explains past Titanic passengers. They would want to know how well the model predicts future or unseen passengers. In my opinion, test accuracy is more realistic because it measures performance on new data.

If `acc_test` happened to be higher than `acc_train`, that would not invalidate holdout testing. It could happen by chance if the test set was slightly easier to predict than the training set. The point of holdout testing is not that test accuracy must always be lower, but that it gives a more honest estimate because the test data was not used to train the model.

```{r compare-accuracy}
# Compare training accuracy and test accuracy
accuracy_comparison <- tibble(
  data = c("Training data", "Test data"),
  accuracy = c(acc_train, acc_test)
)

accuracy_comparison
```

### Exercise 3: Cross validation across timelines

### 3.1

This creates 10 folds. Each passenger is assigned to one fold, and each fold will take a turn being the test set.

```{r create-folds}
# Create a cross-validation datset
# Keep only the variables needed for this model
# Then randomly assign pasengers to 10 folds
set.seed(471)

titanic_cv <- titanic3 %>%
  filter(!is.na(survived), !is.na(sex), !is.na(pclass)) %>%
  mutate(
    fold = sample(rep(1:10, length.out = n()))
  )

# Check the folds
titanic_cv %>%
  count(fold)
```

### 3.2 

```{r fit-evaluate-folds}
# Create an empty results data frame
cv_results <- data.frame(
  fold = sort(unique(titanic_cv$fold)),
  accuracy = NA_real_
)

# Loop through each fold
for (j in cv_results$fold) {
  
  # Use all folds exept j as training data
  train_j <- titanic_cv %>%
    filter(fold != j)
  
  # Use fold j as test data
  test_j <- titanic_cv %>%
    filter(fold == j)
  
  # Fit the model on the trainng folds
  m_j <- glm(
    survived ~ sex + pclass,
    data = train_j,
    family = binomial
  )
  
  # Predict survival probablities for the test fold
  p_j <- predict(m_j, newdata = test_j, type = "response")
  
  # Convert probabilities into class predictions using 0.5 cutoff
  yhat_j <- ifelse(p_j > 0.5, 1, 0)
  
  # Compute accuracy for this fold
  cv_results$accuracy[cv_results$fold == j] <- mean(
    yhat_j == test_j$survived,
    na.rm = TRUE
  )
}

# View fold accuracies
cv_results
```

### 3.3

```{r summarize-cv}
# Summarize cross-validation accuracy
cv_summary <- cv_results %>%
  summarize(
    cv_mean = mean(accuracy),
    cv_sd = sd(accuracy),
    cv_min = min(accuracy),
    cv_max = max(accuracy)
  )

cv_summary
```

### 3.4

```{r visualize-folds}
# Plot accuracy for each fold
ggplot(cv_results, aes(x = factor(fold), y = accuracy)) +
  geom_col() +
  geom_hline(
    yintercept = mean(cv_results$accuracy),
    linetype = "dashed"
  ) +
  labs(
    title = "10-Fold Cross Validation Accuracy",
    subtitle = "Dashed line shows the mean accuracy across folds",
    x = "Fold",
    y = "Accuracy"
  ) +
  theme_minimal()
```

### 3.5

The cross-validation mean is usually lower than the apparent accuracy because cross validation tests the model on passengers that were held out from training. I think this makes it a more honest estimate than apparent acuracy. 

The cross-validation standard deviation tells us how stable the model is across different folds. A small standard deviation means the model performs similarly across folds. A larger standard deviation would mean performance depends more on which passengers happen to be in the test fold.

If Lloyd’s asked for a single performance estimate, I would report the cross-validation mean rather than the one-time test accuracy. The test accuracy is based on just one split, while cross validation averages across multiple splits. I think that makes the cross-validation mean more stable and reliable.

### Exercise 4: When the cutoff is a policy decision

### 4.1

```{r cutoffs}
# Try three possible cutoffs
cutoffs <- c(0.3, 0.5, 0.7)

# Create an empty results data frame
cutoff_results <- data.frame(
  cutoff = cutoffs,
  accuracy = NA_real_
)

# Calculate test accuracy for each cutoff
for (i in seq_along(cutoffs)) {
  
  c0 <- cutoffs[i]
  
  # Predict survival if probability is greater than the cutoff
  yhat_c <- ifelse(p_test > c0, 1, 0)
  
  # Store accuracy
  cutoff_results$accuracy[i] <- mean(
    yhat_c == titanic_test_survival$survived,
    na.rm = TRUE
  )
}

cutoff_results
```

### 4.2

```{r cutoff-plot}
# Try a wider range of cutoffs
cutoffs_fine <- seq(0.1, 0.9, by = 0.05)

# Create an empty results data frame
cutoff_results_fine <- data.frame(
  cutoff = cutoffs_fine,
  accuracy = NA_real_
)

# Calculate accuracy for each cutoff
for (i in seq_along(cutoffs_fine)) {
  
  c0 <- cutoffs_fine[i]
  
  # Convert probabilities into predicted classes using this cutoff
  yhat_c <- ifelse(p_test > c0, 1, 0)
  
  # Store accuracy
  cutoff_results_fine$accuracy[i] <- mean(
    yhat_c == titanic_test_survival$survived,
    na.rm = TRUE
  )
}

# Plot accuracy across cutoffs
ggplot(cutoff_results_fine, aes(x = cutoff, y = accuracy)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Test Accuracy Across Probability Cutoffs",
    x = "Probability cutoff",
    y = "Test accuracy"
  ) +
  theme_minimal()
```

### 4.3

The cutoff that maximized accuracy was **0.65**, with an accuracy of about **0.787**.

Accuracy might be a poor criterion for selecting a cutoff in underwriting because not all mistakes have the same cost. A false positive would mean predicting that a passenger survives when they actually do not survive. 

  - For Lloyd’s, this could be very expensive if the company offers a cheaper policy to       someone who actually dies. A false negative would mean predicting that a passenger        does not survive when they actually does survive. That mistake also matters, but it       may have a different financial consequence.

In my opinion, Lloyd’s should not choose the cutoff based only on accuracy. I would want to use an **expected cost** metric, where false positives and false negatives can be given different costs. Sensitivity and specificity would also be useful because they show whether the model is better at identifying survivors or non-survivors.
 
```{r best-cutoff}
# Find the cutoff with the highest accuracy
cutoff_results_fine %>%
  arrange(desc(accuracy)) %>%
  slice(1)
```

